---
title: "Index Investments (IQ)"
author: "Brian Maja"
date: "30/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
   
  
```{r , include=FALSE}
library(dplyr) 
library(tidyverse) 
library(readxl)  
library(rhandsontable) 
library(purrr) 
library(plotly)
library(scipub)
```


__Source data Preview__
  
```{r}
# souce file
TOP40 = read.csv('TOP40.csv')

# rhandsontable to preview tables
rhandsontable(TOP40,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 
```

### __Task 1__  

Number of rows in the dataset    
  
```{r} 
NROW(TOP40)
```
  
Number of variables (excluding date and Ticker)  
  
```{r} 
data_output = TOP40 %>%
              select(!c(date,Ticker)) #exclude date and Ticker columns
NCOL(data_output)
```
  
Percentage of values for each variable (excluding date and Ticker) the are missing/not available  
    
```{r}
 
sample_list = TOP40 %>%
              select(!c(date,Ticker))%>%  
              map(~ mean(is.na(.))) #use map to apply function mean NA of all column list
 
max_len = max(lengths(sample_list))
df =  purrr::map_df(sample_list, ~ c(., rep('', max_len - length(.)))) #maps the list to dataframe
 
  rhandsontable(df,height = "10vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)  
```
  
### __Task 2__  
   
Calculate and plot the number of stocks available for analysis over all dates  
  
Table output  
  
```{r}
#confirm both Date and ticker have same number of observations
NROW(TOP40$Ticker)
NROW(TOP40$date)

data_output_1 = TOP40 %>%
              group_by(date) %>% #group by date
              summarise(n = n()) #count frequency

rhandsontable(data_output_1,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)  
```
  
Plot output  
  
```{r}
  
plot_output_1 = ggplot(data_output_1, aes(x = as.Date(date), y = n))+ 
                geom_line( colour = 'blue' )+ 
                ylab("Number of stocks available for analysis")+
                xlab("Date")

ggplotly(plot_output_1)

```
  

### __Task 3__    
  
For the month of June 2002, which are the 10 stocks with the highest ROE

```{r}
 
highest_ROE_output = TOP40 %>% 
                      subset(date == "2002-06-30") %>% #subset data to June 2002
                      slice_max(ROE, n = 10) #pick top 10 based on ROE

rhandsontable(highest_ROE_output,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 

```
  
### __Task 4__   
  
Calculate and and plot the average monthly value of X3MAvgDailyValueTradedUSD over all dates.
(X3MAvgDailyValueTradedUSD is the average daily value traded over the previous 3 months for each
stock)
   
__Table output__    
   
```{r}
X3MAvgDailyValueTradedUSD_data = TOP40 %>% 
                                 group_by(date) %>%  
                                 summarise(X3MAvgDailyValueTradedUSD = mean(X3MAvgDailyValueTradedUSD)) #average of the X3MAvgDailyValueTradedUSD

rhandsontable(X3MAvgDailyValueTradedUSD_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 
```
  
__Plot output__    
  
```{r}
 
X3MAvgDailyValueTradedUSD_plot <- plot_ly(X3MAvgDailyValueTradedUSD_data,x = ~as.Date(date), y = ~X3MAvgDailyValueTradedUSD, type = 'scatter', 
               mode = 'lines',fill = 'tozeroy')%>% 
  layout(showlegend = F, title = 'The average daily value traded over the previous 3 months',
         xaxis = list(title = 'Date'),yaxis = list(title = '3-Months Average Daily Value Traded in USD'),
         plot_bgcolor='#e5ecf6')

X3MAvgDailyValueTradedUSD_plot
```
  
Commentary:  
  
There has been an increase in the liquidity of the top40 traded stocks over the years.  
   
### __Task 5__   
  
Calculate and plot the average monthly value of ROE over all dates.
  
__Table output__  
  
```{r}
ROE_data = TOP40 %>% 
            group_by(date) %>%  
            summarise(ROE = mean(ROE,na.rm = T)) #average of the ROE

rhandsontable(ROE_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 
```
  
__Plot output__    
  
```{r} 
 
ROE_plot <- plot_ly(ROE_data,x = ~as.Date(date), y = ~ROE, type = 'scatter', mode = 'lines')%>% 
              layout(showlegend = F, title = 'The average monthly value of ROE',
              xaxis = list(title = 'Date'),yaxis = list(title = 'Average monthly value of ROE'),
              plot_bgcolor='#e5ecf6')

ROE_plot

```

  
Commentary:  
  
There has been a steady decline in the Return on Equity/ROE since the year 2014, this basically means that the top40 have generally not been performing well.  
    
### __Task 6__   
  
Confirm that the Market Cap weights (MarketCapUSD) for each date, over all dates sum to 100%  
  
```{r}
MarketCapUSD_data = TOP40 %>% 
                    group_by(date) %>%  
                    summarise(MarketCapUSD = sum(MarketCapUSD,na.rm = T)) #sum excluding NA

rhandsontable(MarketCapUSD_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 
```
  
__Commentary:__    
  
We can see from above that the Market Cap weights does not sum to 100%, we will proceed with normalizing it to 100% below  
  
```{r}
MarketCapUSD_data_norm = TOP40 %>% 
                          group_by(date) %>%  
                          summarise(MarketCapUSD = sum(MarketCapUSD,na.rm = T)) %>%
                          mutate(MarketCapUSD_Normalised = MarketCapUSD/max(MarketCapUSD) * 100)

rhandsontable(MarketCapUSD_data_norm,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)  
``` 
  
__Normalised Market Cap plot__  
  
  
```{r} 
 
fig <- plot_ly(MarketCapUSD_data_norm,x = ~as.Date(date), y = ~MarketCapUSD_Normalised, type = 'scatter', 
               mode = 'lines')%>% 
  layout(showlegend = F, title = 'The Market Cap weights',
         xaxis = list(title = 'Date'),yaxis = list(title = 'Normalised Market Cap weights in %'),
         plot_bgcolor='#e5ecf6')

fig

```  
    
### __Task 7__  
   
   
Instead of using the average weight (in other words, equally weighted), please use the market cap
weights you calculated in Task 6 to calculate the Market Cap Weighted ROE for the TOP40 for each
month, and plot the result for all dates  

Since the question was not explicit as to which weighting to use, I used the not normalised MarketCapUSD
 
__Table output__   
  
```{r}
Market_Cap_Weighted_ROE_data = TOP40 %>% 
                                group_by(date) %>%  
                                summarise(MarketCapUSD_ROE = sum(MarketCapUSD*ROE,na.rm = T))  

rhandsontable(Market_Cap_Weighted_ROE_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T) 
```    

__Plot output__

```{r} 

Market_Cap_Weighted_ROE_plot <- plot_ly(Market_Cap_Weighted_ROE_data,x = ~as.Date(date), y = ~MarketCapUSD_ROE, type = 'scatter', 
               mode = 'lines')%>% 
                layout(showlegend = F, title = 'The Market Cap Weighted ROE for the TOP40',
                  xaxis = list(title = 'Date'),yaxis = list(title = 'Market Cap Weighted ROE for the TOP40'),
                  plot_bgcolor='#e5ecf6')

Market_Cap_Weighted_ROE_plot

```   
   
### __Task 8__  
  
Instead of using the average weight (in other words, equally weighted), please use the market cap
weights you calculated in Task 6 to calculate the Market Cap Weighted Return over the next month
(RON1m) for the TOP40 for each month, and plot the result for all dates  
   
__Table output__       
   
```{r}
marketcapusd_ron1m_data = TOP40 %>% 
                          group_by(date) %>% 
                          summarise(MarketCapUSD_RON1m = sum(MarketCapUSD*RON1m,na.rm = T))  
 
rhandsontable(marketcapusd_ron1m_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)
```  
  
__Plot output__         
   
```{r}  

fig <- plot_ly(marketcapusd_ron1m_data,x = ~as.Date(date), y = ~MarketCapUSD_RON1m, type = 'scatter', 
               mode = 'lines')%>% 
               layout(showlegend = F, title = 'The Market Cap Weighted Return over the next month (RON1m) for the TOP40',
                xaxis = list(title = 'Date'),yaxis = list(title = 'Market Cap Weighted Return over the next month (RON1m) '),
                 plot_bgcolor='#e5ecf6')

fig
 

``` 

### __Task 8b (Bonus Question)__  
    
Using the result from Step 8, plot the value of R1 invested at the start of the analysis in the Market
Cap Weighted portfolio, over time.  

__Table output__  
  
```{r}
R1investmentMWP_data = TOP40 %>% 
                        group_by(date) %>%  
                        summarise(MarketCapUSD_RON1m = sum(MarketCapUSD*RON1m,na.rm = T))%>%  
                          mutate(MarketCapUSD_RON1m_Overtime = 1+cumsum(MarketCapUSD_RON1m))

rhandsontable(R1investmentMWP_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)
```  
  
__Table output__  
  
  
R1 invested at the start of 2002 will be R2.9251208 at the end of the duration
  
```{r}
fig <- plot_ly(R1investmentMWP_data,x = ~as.Date(date), y = ~MarketCapUSD_RON1m_Overtime, type = 'scatter', 
               mode = 'lines')%>% 
               layout(showlegend = F, title = 'R1 invested in the Market Cap Weighted Portfolio',
                xaxis = list(title = 'Date'),yaxis = list(title = 'Market Cap Weighted Return over the next month (RON1m) '),
                 plot_bgcolor='#e5ecf6')

fig
```  
    
### __Task 9__    
    
Grouped on a monthly basis and using the ROE column, calculate the Z-Score for every stock at each
point in time.  
Replace missing values with zeroes  
The name for the resulting column should be “Z_ROE”  
Winsorise outlier Z-scores at 3  
  
__Definition:__ Z-score, sometimes called standard score, is a measurement of how many standard deviations a point is away from the mean of its data set.  
  
```{r} 
z_score_data = TOP40 %>%  
                mutate(ROE = replace_na(ROE,0))%>% #remove NA in ROE
                group_by(date)%>%
                mutate(Z_ROE = (ROE-mean(ROE))/sd(ROE))%>% #calculate Z-ROE
                mutate(Z_ROE_winsor = winsorZ(Z_ROE, zbound = 3)) %>% #Winsorise at 3
                select(date, Ticker, ROE, Z_ROE, Z_ROE_winsor) #Subset data

rhandsontable(z_score_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)

```  
  
### __Task 10__  
  
Plot the Z-score for both ‘Anglo American PLC’ and ‘MTN Group Ltd’ through time  
  
```{r}
z_scoreplot =  z_score_data %>%
                filter(Ticker == 'MTN Group Ltd' | Ticker == 'Anglo American PLC')%>%
                select(date, Ticker,  Z_ROE)
 
z_scoreplot = ggplot(z_scoreplot, aes(x = as.Date(date), y = Z_ROE)) + 
              geom_line(aes(color = Ticker, linetype = Ticker)) + ylab("Z-score") + xlab("Date")+
              scale_color_manual(values = c("darkred", "darkblue"))

ggplotly(z_scoreplot)
```  
  
### __Task 11__   
  
Narrow the dataset to only the following columns  
- date  
- Ticker  
- MOMENTUM  
- QUALITY  
- RON1m  
  
Replace missing values with Zero  
Calculate the Rank (highest value = 1) per month for every stock for the 3 variables  
  

```{r}
ranked_data =  TOP40 %>%   
                select(date, Ticker, MOMENTUM, QUALITY, RON1m)%>%
                mutate(MOMENTUM = replace_na(MOMENTUM,0))%>%
                mutate(QUALITY = replace_na(QUALITY,0))%>%
                mutate(RON1m = replace_na(RON1m,0))%>%
                arrange(MOMENTUM, QUALITY, RON1m) #rank data by 3 variables  

rhandsontable(ranked_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)

```  
  


  
### __Task 12__  
  
Calculate the correlation between the ranked variables (QUALITY and MOMENTUM) and ranked
RON1m over all dates  
    
  
```{r}
  
correlation_data =  ranked_data %>%    
                    mutate(Cor_Quality_RON1m = cor(QUALITY, RON1m, method ='spearman'))%>%
                    mutate(Cor_Momentum_RON1m = cor(MOMENTUM, RON1m, method ='spearman'))%>%
                    select(date, Ticker, Cor_Quality_RON1m,Cor_Momentum_RON1m) 
 
rhandsontable(correlation_data,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)
```  
  
Commentary:  
  
Information Coefficient evaluates the skills of the portfolio manager at how closely their forecasts are to the actuals 
IC values range from -1 to 1 with -1 indicating that the manager fails at making correct predictions whilst 1 is the converse.  
IC value of 0 means that there's no linear relationship  
Since the value IC value is 0.02917685 and 0.0464988 this shows that the Manager is generally performing better  
  
Why would we use the Spearman (ranked) correlation as opposed to the more usual Pearson correlation  
  
We use the Spearman (ranked) correlation method because the variables MOMENTUM, QUALITY, RON1m are not normaly distributed thus we have to use non-parametric correlation  
  
We can test for normality using shapiro-wilk test and qqplot  
  
 

  
### __Task 12b (Bonus)__  
  
```{r}
  
variables = c("QUALITY","MOMENTUM")
Accumulated_IC = c(sum(correlation_data$Cor_Quality_RON1m),sum(correlation_data$Cor_Momentum_RON1m))
df_comparison = data.frame(variables,Accumulated_IC) #create comparison dataframe
 
correlation_plot = ggplot(data=df_comparison, aes(x=variables , y=Accumulated_IC)) +
  geom_bar(stat="identity", fill="steelblue")+ ylab("Accumulated Information Coefficient Over Time") + xlab("Portfolios")+
  theme_minimal()

ggplotly(correlation_plot)

```  
  
From the accumulated information coefficient above its clear that MOMENTUM has the best ability to forecast future returns  
  
### __Task 13__  
  
Using the IC data for the two variables, calculate a t-statistic for each
  
```{r}
#reference: https://stats.stackexchange.com/questions/17281/what-does-t-statistics-of-information-coefficient-indicate
#results confirmed with https://www.researchgate.net/post/Can-we-test-the-significance-of-spearman-correlation-using-t-test

#statistic for the 1-sample t-test
 sqrt(NROW(correlation_data))*correlation_data$Cor_Quality_RON1m[1]
 sqrt(NROW(correlation_data))*correlation_data$Cor_Momentum_RON1m[1]

```  
   
T test values for QUALITY and MOMENTUM are 2.622673 and 4.179722 respectively  
  
Comment on and infer a confidence level for using each of the two variables to invest client money
with.

```{r}
  
cor.test(ranked_data$MOMENTUM, ranked_data$RON1m,
         method = "spearman", exact = F)
#----------------
cor.test(ranked_data$QUALITY, ranked_data$RON1m,
         method = "spearman", exact = F)
```  
  
__MOMENTUM portfolio analysis:__  
  
From the Spearman correlation coefficient rho is 0.0464988, since the coefficient value is positive this means that theres a
positive correlation between the 2 variables MOMENTUM and RON1m. The p value is very small below the alpha level of 0.05, as a result we will  
reject the null and accept the alternative in that theres a significant positive correlation between MOMENTUM and RO1m
  
__QUALITY portfolio analysis:__  
  
From the Spearman correlation coefficient rho is 0.02917685, since the coefficient value is positive this means that theres a   
positive correlation between the 2 variables QUALITY and RON1m. The p value however is large just above the alpha level of 0.05, as a result we fail to
reject the null in that theres no a significant positive correlation between MOMENTUM and RO1m.
  
As a result, for investing clients money we are statistically confident that MOMENTUM portfolio will the way to go at a 5% leve of significance.  
  
### __Task 14__  
  
Build an equally weighted portfolio every month of the 10 stocks with the highest MOMENTUM  
  
  
```{r}
stock_picks = ranked_data%>%
              group_by(date)%>%
              slice_max(MOMENTUM, n=10) #select top 10 stocks with highest momentum per month

stock_picks_RON1m = stock_picks%>%
                    group_by(date)%>%
                    summarise(RON1m = sum(RON1m))

rhandsontable(stock_picks_RON1m,height = "30vh",
              readOnly = TRUE,search = TRUE)%>%
              hot_cols(columnSorting = TRUE,highlightCol = TRUE, highlightRow = TRUE,
              manualColumnResize = T)
```  
  
### __Task 14b (Bonus Question)__    
  
Using the result from Step 14, plot the value of R1 invested at the start of the analysis in the 10 stock
MOMENTUM portfolio, over time.  
   
  
```{r}

R1_stock_picks_RON1m = stock_picks %>% 
                        group_by(date) %>% 
                        summarise(Momentum_RON1m = sum(MOMENTUM*RON1m,na.rm = T))%>%  # momentum weighting
                          mutate(MOMENTUM_RON1m_Overtime = 1+cumsum(Momentum_RON1m)) # cummulative sum on 1 Rands investment
 
R1_stock_picks_RON1m_plot = ggplot(R1_stock_picks_RON1m, aes(x=as.Date(date), y=MOMENTUM_RON1m_Overtime)) +
                            ylab("R1 invested on the MOMENTUM portfolio") + xlab("Date")+
                              geom_line(colour ='blue')
 
ggplotly(R1_stock_picks_RON1m_plot)
```  
  
### __Task 15 (Bonus Question)__    
  
Using the result from Step 8b and 14b, graphically compare the performance of the 10 stock MOMENTUM portfolio to the Market Cap weighted portfolio.

```{r}
 
#rename columns to avoid join conflict
colnames(R1investmentMWP_data) = c('date','MarketCapweight_RON1m','MarketCapweight_Overtime')
 

 
joined_df = left_join(R1investmentMWP_data,R1_stock_picks_RON1m) #join 2 dataframes

joined_plot = ggplot(joined_df, aes(as.Date(date))) + 
  geom_line(aes(y =  MarketCapweight_Overtime, colour = "Market Cap weighted portfolio")) + 
  geom_line(aes(y =  MOMENTUM_RON1m_Overtime, colour = "MOMENTUM portfolio"))+
  ylab("Value in Rands") + xlab("Date") 
ggplotly(joined_plot)
 
```  